---
title: "Notebook 09: Metrics: G-Score"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r message=FALSE}
library(tidyverse)
library(forcats)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
```

## Amazon Products Again

For a final time, let's look at the spam text message data.

```{r, message=FALSE}
set.seed(1)

amazon <- read_csv("data/amazon_product_class.csv") %>%
  mutate(train_id = if_else(runif(n()) < 0.6, "train", "valid"))
token <- read_csv("data/amazon_product_class_token.csv.gz")
```

We will see today a simple metric that we can use in parallel to the
coefficients from a penalized linear regression.

### Dunn-Likelihood, G-squared

Given a textual feature (such as a word count) and a variable that groups a
corpus into two groups, the Dunn-likelihood, or G-squared, metric indicates how
strongly the feature is associated with one group over the other. To compute
the measurement, we use the function `sm_dunn_ll`, providing the name of the
grouping variable and the feature variable. Note that this often requires
(1) joining together the token data set and the text data set and (2) grouping
together categories. Be careful to drop off the text column so as not to create
many large copies of the corpus.

```{r}
left_join(token, select(amazon, doc_id, category), by = "doc_id") %>%
  mutate(category = if_else(category == "book", category, "other")) %>%
  sm_dunn_ll(group_name = "category", token_name = "lemma") %>%
  arrange(dunn)
```

The score can be used equally as well with other features, such as the
universal part of speech codes:

```{r}
left_join(token, select(amazon, doc_id, category), by = "doc_id") %>%
  mutate(category = if_else(category == "food", category, "other")) %>%
  sm_dunn_ll(group_name = "category", token_name = "xpos") %>%
  arrange(desc(dunn))
```

And again, this is similar to the patterns we saw looking at the coefficients
of a penalized regression model.

What are the relative benefits and drawbacks of the G-squared metric relative
to running a penalized regression model? The G-squared metric is useful because
it provides an absolute number that does not depend on the other features in the
model or the tuning parameters (alpha and lambda) used in the model. It can be
safely compared across corpora even when the length of documents or frequency
of terms is very different between the two. One downside is that it does not
provide an overall accuracy or confusion matrix that allows us to see how
useful a feature is in distinguishing two categories. It also does not allow
us to include additional covariates. Perhaps the biggest downside is that
the G-squared does not work in the case of more than two categories.

Generally, I recommend using the two in tandem. I typically start with the
predictive model and then use the G-squared values to indicate the strength
of a relationship once I have identified interesting terms. Also, I find the
G-squared value particularly useful when looking at the frequency of POS tags
or dependency relationships between two groups.
