---
title: "Notebook 14: Named Entity Recognition"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r message=FALSE}
library(tidyverse)
library(forcats)
library(ggrepel)
library(smodels)
library(cleanNLP)
library(magrittr)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
```

## Wikipedia NER

For our final set of notes, we will once again look at the Wikipedia data set
created by following the links from the Coffee and Tea pages.

```{r, message = FALSE}
wiki <- read_csv(file.path("data", "wiki_coffee_tea.csv"))
```

We have been using the results from parsing text throughout the semester. We
have used the tokens table produced by the `cnlp_annotate` function, but when
using the spaCy backend, there is another table of annotations that can
sometimes by very useful called `entity`. For reference, this is how to extract
it from the output:

```{r, eval=FALSE}
cnlp_init_spacy("en_core_web_sm")
anno <- cnlp_annotate(wiki)
token <- anno$token
ner <- anno$entity
```

```{r, echo=FALSE, message=FALSE}
token <- read_csv(file.path("data", "wiki_coffee_tea_token.csv.gz"))
ner <- read_csv(file.path("data", "wiki_coffee_tea_entity.csv.gz"))
```

The process of finding *entities* is called Named Entity Recognition (NER). It
is a very active area of natural language processing. This is what the data
detected by the named entity recognition process looks like:

```{r}
ner
```

The algorithm has identified one or more tokens corresponding to known entities,
such as dates, people, or locations. Here are some of the different entity types
included in the data:

```{r}
ner %>%
  group_by(entity_type) %>%
  summarize(sm_count()) %>%
  arrange(desc(count))
```

More details about the entity types can be found in the
[spaCy documentation](https://spacy.io/api/annotation#named-entities).

Entity detection has a large number of useful applications; for us we will 
mostly use it to describe documents. This might be very useful for the final
project, which works with one of two different news corpora. For example, we
can detect the GPE (countries, cities, states) tags that occur most frequently
(and at least 20 times) in each document:

```{r}
ner %>%
  filter(entity_type == "GPE") %>%
  group_by(doc_id, entity) %>%
  summarize(sm_count()) %>%
  arrange(doc_id, desc(count)) %>%
  slice_head(n = 1) %>%
  filter(count >= 20) %>%
  mutate(lab = paste(doc_id, entity, sep = " => ")) %>%
  use_series(lab)
```

Or, similarly, languages that are mentioned at least 4 times

```{r}
ner %>%
  filter(entity_type == "LANGUAGE") %>%
  group_by(doc_id, entity) %>%
  summarize(sm_count()) %>%
  arrange(doc_id, desc(count)) %>%
  slice_head(n = 1) %>%
  filter(count >= 4) %>%
  mutate(lab = paste(doc_id, entity, sep = " => ")) %>%
  use_series(lab)
```

Dates can be useful too, though some are more useful that others:

```{r}
ner %>%
  filter(entity_type == "DATE") %>%
  group_by(doc_id, entity) %>%
  summarize(sm_count()) %>%
  arrange(doc_id, desc(count)) %>%
  slice_head(n = 1) %>%
  filter(count > 10) %>%
  mutate(lab = paste(doc_id, entity, sep = " => ")) %>%
  use_series(lab)
```

I do not have much more to say about these other than that they can be useful
for specific tasks. For example, you can use them to find characters in a novel
or for automatically linking pages to secondary sources. This later step
requires an additional step of entity linking, which is not currently supported
in the production version of **cleanNLP**, but should be soon.
