---
title: "Notebook 02: Creating the Model Matrix"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r message=FALSE}
library(tidyverse)
library(forcats)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
```

## Predicting Bike Usage

Today, we will look at a data set of bike sharing usage in Washington, DC.
Our task is to predict how many bikes are rented each day. As we saw in the
previous set of notes, it will be useful to split the data set into training
and validation sets. We will do this right as the data is read into R.

```{r, message = FALSE}
set.seed(1)

bikes <- read_csv("data/bikes.csv") %>%
  mutate(train_id = if_else(runif(n()) < 0.6, "train", "valid"))
bikes
```

### Non-linear relationships

The data comes from two different years. There was an overall increase in usage
during the second year, so it is useful to start by looking at just the second
year (we will see how to incorporate both together in just a moment). Here is a
plot that investigates the relationship between the maximum daily temperature
(degrees Celsius) and the number of bikes that are rented.

```{r}
bikes %>%
  filter(year == "two") %>%
  ggplot(aes(temp, count)) +
    geom_point()
```

Generally, people are more willing to rent bikes when the weather is warmer.
However, the usage drops a bit when the temperatures are too high. The peak
usage visually seems to be right around 20-25°, which is right around the
optimal temperature for a bike ride.

We will start by building a linear model between the temperature and the bike
usage in year two. Rather than worrying about the RMSE, we will start by just
visually inspecting the results.

```{r}
model <- bikes %>%
  filter(year == "two") %>%
  filter(train_id == "train") %>%
  lm(count ~ temp, data = .)

bikes %>%
  filter(year == "two") %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point() +
    geom_line(aes(y = score_pred), color = "orange", size = 1.5)
```

The best fit line seems to over-predict usage at very low temperatures and
very high temperatures. We already noted that the relationship is not a linear
one, so it is not surprising that linear regression line fails to match the
data for all temperature values.

You might think that we need an entirely different kind of model to deal with
this data. After all, we have a non-linear relationship, so how can we use a
linear model? It turns out, helpfully, that this is in fact not the case because
linear models are only constrained to be linear in the sense that the unknown
coefficients must be linearly related to the output. However, we can create
additional predictor variables to fit high order relationships. For example,
we can fit the following model to predict a quadratic relationship between
x and y:

$$ Y_i = \alpha + \beta_1 X_{i} + \beta_2 X_{i}^2 + \epsilon_i  $$

In other words, we have converted a non-linear model with one variable into a
linear model with two variables. Let's see how this could work in R by creating
a temporary variable of the squared temperature:

```{r}
model <- bikes %>%
  filter(year == "two") %>%
  filter(train_id == "train") %>%
  mutate(temp_sq = temp^2) %>%
  lm(count ~ temp + temp_sq, data = .)

bikes %>%
  filter(year == "two") %>%
  mutate(temp_sq = temp^2) %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point() +
    geom_line(aes(y = score_pred), color = "orange", size = 1.5)
```

Now, the model fits the data much more accurately at the extreme temperature
values. An alternative way of fitting the same model without having to worry
about manually constructing a new variable of squares is through the `poly()`
function. It does all of the work for us; while in theory it produces the same
fit as the code above, it does so using a more numerically stable algorithm.

```{r}
model <- bikes %>%
  filter(year == "two") %>%
  filter(train_id == "train") %>%
  lm(count ~ poly(temp, degree = 2), data = .)

bikes %>%
  filter(year == "two") %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point() +
    geom_line(aes(y = score_pred), color = "orange", size = 1.5)
```

The `poly` function can be used to fit higher-order interactions by increasing
the degree of the polynomial.

```{r}
model <- bikes %>%
  filter(year == "two") %>%
  filter(train_id == "train") %>%
  lm(count ~ poly(temp, degree = 5), data = .)

bikes %>%
  filter(year == "two") %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point() +
    geom_line(aes(y = score_pred), color = "orange", size = 1.5)
```

Here, the fifth-order polynomial does a better job of capturing the extreme
tails of the data: a smaller slope for very low temperatures and a sharper
one for high temperatures.

### Indicator variables

Now, let's try to get a handle on the two different years of the data. Drawing
the temperature-count relationship with both years shows the large increase
in usage from year one to year two.

```{r}
bikes %>%
  ggplot(aes(temp, count)) +
    geom_point(aes(color = year), alpha = 0.5)
```

We can also model this relationship with a linear model by constructing another
new variable called an *indicator variable*. This variable takes on only two
values: 0 or 1. It is defined to be one in year two and zero otherwise. If we
call this variable Z, we can think of our linear model being:

$$ Y_i = \alpha + \beta_1 X_{i} + \beta_3 Z_{i} + \epsilon_i  $$

Constructing this variable manually, we can create a model in R with the
following:

```{r}
bikes %>%
  mutate(yeartwo = if_else(year == "two", 1, 0)) %>%
  filter(train_id == "train") %>%
  lm(count ~ temp + yeartwo, data = .) %>%
  summary()
```

The coefficient for the `yeartwo` variable (2103.58) gives the increase in bikes
rented on average between year one and year two after accounting for the
temperature variable. As with the polynomial expansion in the previous section,
we can let R handle the creation of the indicator variable by just adding the
year variable in to the R formula.

```{r}
bikes %>%
  filter(train_id == "train") %>%
  lm(count ~ temp + year, data = .) %>%
  summary()
```

Plotting the model now shows that the predictions are given by two parellel
lines.

```{r}
model <- bikes %>%
  filter(train_id == "train") %>%
  lm(count ~ temp + year, data = .)

bikes %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point(aes(color = year), alpha = 0.5) +
    geom_line(aes(y = score_pred, color = year), size = 1.5)
```

We can, of course, combine this with the polynomial expansion to have two
polynomials with different intercepts.

```{r}
model <- bikes %>%
  filter(train_id == "train") %>%
  lm(count ~ poly(temp, 5) + year, data = .)

bikes %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point(aes(color = year), alpha = 0.5) +
    geom_line(aes(y = score_pred, color = year), size = 1.5)
```

Sometimes it can even be useful to have different slopes for the two lines. To
do this, we multiply the two variables together in the model formula.

```{r}
model <- bikes %>%
  filter(train_id == "train") %>%
  lm(count ~ temp * year, data = .)

bikes %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point(aes(color = year), alpha = 0.5) +
    geom_line(aes(y = score_pred, color = year), size = 1.5)
```

### Indicator variables: Three or more categories

Finally, the approach above works equally well when we have a variable with
more than two categories. The R syntax and the geometric interpretation
extend as you would expect. For example, if we use the weather variable,
which has three categories (Dry, Rain, and Snow) we now have three parallel
curves:

```{r}
model <- bikes %>%
  filter(train_id == "train") %>%
  lm(count ~ poly(temp, 5) + weather, data = .)

bikes %>%
  mutate(score_pred = predict(model, newdata = .)) %>%
  ggplot(aes(temp, count)) +
    geom_point(aes(color = weather), alpha = 0.2) +
    geom_line(aes(y = score_pred, color = weather), size = 1.5)
```

Unsurprisingly, rain decreases usage a bit and snow decreases significantly.
Under the hood, R is created two indicator variables, one an indicator for Rain
and the other for Snow. The details are important if you want to do statistical
inferences, but the visualization is sufficent for understanding the concept
from a predictive perspective.

## Linear Models with Matrices

### Matrix Formulation

We have been working with multivariate linear models with several variables,
though I have only ever written the formal equation in the case where there
are two explanatory variables. In general, multivariate regression represents
the following model:

$$y_i = x_{1,i} \beta_1 + x_{2,i} \beta_2 + \cdots + x_{1,p} \beta_p + \epsilon_i$$

For simplicity, we won't include an explicit intercept term in
the model. If we want one, we will just make the first variable $x_{1,i}$
equal to one for every value of i.

The statistical estimation problem is to estimate the p
components of the multivariate vector beta.

Using the notation of matrix multiplication, we can write the linear model
simultaneously for all observations:

$$ \left(\begin{array}{c}y_1\\ y_2\\ \vdots\\ y_n\end{array}\right) =
  \left(\begin{array}{cccc}x_{1,1}&x_{2,1}&\cdots&x_{p,1}\\
                           x_{1,2}&\ddots&&x_{p,2}\\
                           \vdots&&\ddots&\vdots\\
                           x_{1,n}&x_{2,n}&\cdots&x_{p,n}\\\end{array}\right)
  \left(\begin{array}{c}\beta_1\\ \beta_2\\ \vdots\\ \beta_p\end{array}\right) +
  \left(\begin{array}{c}\epsilon_1\\ \epsilon_2\\ \vdots\\ \epsilon_n\end{array}\right) $$


Which can be compactly written as:

$$ y = X \beta + \epsilon $$

The matrix X is known as the **design matrix** or **model matrix**.

## Fitting Linear Models with Matricies

Note that the bikes data, like every other data set we have used,
is not a matrix. It is something that R calls a **data frame**:

```{r}
class(bikes)
```

While both matrices and data frames have data organized in rows and
columns, matrices force all of the data to be numeric (its actually
more complicated than this in R, but just go along with it for now).
A data frame on the other hand can have different variable types in
each column.

The formula interface to `lm` is incredibly useful when using
categorical variables or when processing numeric variables by special
functions such as `poly`. The functions `model.frame` and `model.matrix`
allow us to compute the model matrix from the formula interface. In fact, the
`lm` function calls these to convert our inputs into a model matrix. The
output of this is then passed to `lm.fit`. It will also, by
default, include an intercept term for us. Here we use it
to build a model matrix as before:

```{r}
mf <- model.frame(count ~ temp +  humidity , data = bikes)
mt <- terms(mf)
y <- model.response(mf)
X <- model.matrix(mt, mf)

head(X)
```

Notice that the intercept has been added for us. What is nice about this
formulation is that we can use commands like `poly` and categorical
variables and have R take care of all the hard work for us:

```{r}
mf <- model.frame(count ~ temp + weather, data = bikes)
mt <- attr(mf, "terms")
y <- model.response(mf)
X <- model.matrix(mt, mf)

head(X)
```

We can then create a specific training set using the following syntax:

```{r}
X_train <- X[bikes$train_id == "train", ]
y_train <- y[bikes$train_id == "train"]
```

We will find that many of the more complex algorithms for fitting models in
R require the user to produce a model matrix before passing the data to the
statistical learning algorithm.

## Using lm.fit

We have seen how to use the `lm` function to quickly fit
linear regression directly from data frames. Now, how do
we actually fit a linear model once we have these matrices?
Next class we will see how to do this directly with matrix
operations. There is an intermediate function that solves the
linear regression problem directly from our matrices called
`lm.fit`. In fact, the `lm` function internally calls this function.
As inputs, it takes just the X matrix and response y. There
is a lot of diagnostic output, but we will take just the
`coef` component, corresponding to the coefficients matrix:

```{r}
beta <- lm.fit(X_train, y_train)$coef
beta
```

We can create predicted values for the whole data set by matrix
multiplying `X` with `beta`:

```{r}
bikes %>%
  mutate(count_pred = X %*% beta) %>%
  select(count, count_pred)
```

Let's verify that this gives the same output as the `lm` function:

```{r}
bikes %>%
  filter(train_id == "train") %>%
  lm(count ~ temp + weather, data = .) %>%
  summary()
```

And, as hoped, it does!

Generally, it is recommended to us the `lm` function directly if your goal is
just to produce a linear regression model. Creating a model matrix is a useful
process to know, however, as we move towards more complex learning methods.

### Direct computation (optional)

Some of you may have taken a class where you derived the analytic form of the
linear regression estimator. It is sometimes shown in a linear algebra class or
advanced econometrics course. We can compactly write the estimator as the
following matrix product:

$$ \hat{\beta} = (X^t X)^{-1} X^t y $$

And sure enough, this is the same solution given by the `lm` and `lm.fit`
functions:

```{r}
solve(crossprod(X_train), crossprod(X_train, y_train))
```

We will explore the use of matrix products later in the semester when we look
at PCA and spectral clustering.
