---
title: "Using R to Manipulate and Visualize Data"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Introduction

My statistical learning course (MATH389) is in many ways a natural extension of
the way  that I teach Introduction to Data Science (MATH289). I recognize that
many students this semester, however, may have taken 289 with a different
instructor in a previous year or have come through one of the alternative
prerequisites. The course should be accessible to anyone with previous
experience writing code to analyze data, provided that one is prepared to put
in a bit of work at the start of the semester to fill in any gaps of knowledge.

These notes provide a starting point for getting up to speed with the specific
material covered in 289. They give an overview of some of the R functions we
will be working with this semester. The notes assume that you have had some
previous exposure to writing code, but no particular exposure to R. For example,
perhaps you have used Python in an introductory CS course of Matlab in a
mathematics course. Even if you have used R in other courses, it is quite
possible that you have not used the specific functions and approaches mentioned
here, so please review these notes carefully.

There is no way that these notes will completely substitute for all of the
detail that we go into in MATH289, Introduction to Data Science. However, they
should give a good baseline from which to understand the code that we will
develop throughout the course. For more information, I recommend that you check
out:

- [MATH289 Course Notes](https://statsmaths.github.io/stat289-f20/index.html)
- [R for Data Science Textbook](https://r4ds.had.co.nz/)

I make references to both of these in several points throughout these notes
as topics arise. I am always happy to answer questions about specific functions
or approaches being used in the course notes.

### Installation

The R programming language and all of the third-party packages that we will
use during the semester is free and open-source. As a class, we will walk
through the steps to install the language and needed components on your machine
during the first course meeting. In the event of difficulties, we have a
subscription to a cloud-based alternative that you are able to make use of.
For now, I recommend just reading and understanding the code described here.
There will be plenty of time to practice the material once the semester begins.

If the entire concept of writing and running scripting code seems foreign to
you, I suggest reading chapters 1 and 4 of the R for Data science book, which
is freely available here:

- [Chapter 1 – Introduction](https://r4ds.had.co.nz/introduction.html)
- [Chapter 4 – Workflow: Basics](https://r4ds.had.co.nz/workflow-basics.html)

In the case that these chapters seem to be moving too fast, which is hopefully
not the case, I recommend contacting me to assess whether 389 is an appropriate
course based on your background.

### Setup

Usually, the first we do when running R code is to load third-party extensions
called *packages*. These provide additional functions that make working with
data easier and more consistent than using build-in function. Here are three
packages that I tend to use in my data analysis work:

```{r message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
```

Next, we typically need to set some parameters that change the default values
for common functions. Here are three that I also typically include, that
change the way that plots, output, and data summaries are performed:

```{r message=FALSE}
theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
```

Typically in class, I will include all of the packages and default settings
that you need for an analysis.

### Loading and Viewing Datasets

After loading the R libraries and setting up our work space, the next step is
to load in some data. Here, we load a csv file into R and store the data set
as an R object named `food`. Notice that R uses the arrow sign (`<-`) to
assign the output of one function to a variable name.

```{r, message = FALSE}
food <- read_csv(file.path("data", "food.csv"))
```

The food data set contains information about various food items, with one row
for each item of food (we call these observations) and one column for each
thing that we know about the food items (we call these features). Running a
line of code with just the name of the data set prints out the first few rows
and columns of the data. Additional feature names are given at the bottom of
the output.

```{r}
food
```

Notice that the each of the columns has a unique name. For this class, we will
use the convention that all column and variable names use a combination of
lower-case letters, underscores, and numbers. This will helps keep our code neat
and easy to follow.

## Manipulating Data

### Data Verbs

A common task in data analysis is manipulating an existing data set. In R, we
will often do this through the use of data verbs. These are functions that
take one version of a data set and return a modified version of the data set.
They always work on a copy of the data. Often we want to apply a sequence of
data verbs one after another. To make this easy, and avoid the need to create
temporary variables, we can use the pipe operator `%>%`, which passes the
output of one line into the first argument of the next.

It is probably more intuitive to see an example of data verbs and the pipe
operator, rather than trying to overly-describe the way it functions. As an
example, here is a chain of operators that first **filters** the data to include
only those rows where the food group various indicates a fruit and then
**arranges* the data from the smallest to the largest value of sodium.

```{r}
food %>%
  filter(food_group == "fruit") %>%
  arrange(sodium)
```

Other common verbs include the **mutate** verb, which creates new variables
as a function of the existing variables, and the **select** verb which selects
a subset of the existing columns. Here, we compute the percentage of calories
that are from fat (there are 9 calories in 1 gram of fat), arranges the
data in descending order by the percentage of calories that are from fat, and
then selects only the relevant variables.

```{r}
food %>%
  mutate(calories_fat_perc = total_fat * 9 / calories * 100) %>%
  arrange(desc(calories_fat_perc)) %>%
  select(item, total_fat, calories, calories_fat_perc)
```

Inside of the filter function you can use a number of different logical
operators, such as `>`, `<`, `>=`, `<=`, `!=` (not equal), `%in%`
(set containment), `&` (and), and `|` (or). And inside of the mutate function,
an array of different mathematical operators can be applied, such as `sqrt()`,
`sin()`, and `abs()`.

### Summarizing Data

Another data manipulation verb that deserves special attention is the
**summarize** command. By default, it summarizes all of the rows of a data set
in a single line by applying *summary functions* to columns of the data. For
example, here is the code to take the average (mean) value of three of the
variables in the food data:

```{r}
food %>%
  summarize(sm_mean(calories), sm_mean(total_fat), sm_mean(sat_fat))
```

Notice that only the new variables are included in the output.

The real power of the summary function comes by grouping the data by one or
more variables prior to running the summary command. The summary function will
then be applied only *within* each unique value of the grouping variable(s),
with one row for each unique value. Here is the code to compute the same
summaries for each food group:

```{r}
food %>%
  group_by(food_group) %>%
  summarize(sm_mean(calories), sm_mean(total_fat), sm_mean(sat_fat))
```

There are number of other summary functions that we will occasionally need,
such as `sm_min()`, `sm_median`, `sm_quantiles()`, `sm_count()`, and `sm_cor()`.
It is also possible to use a format similar to the mutate function, where the
summary function and new variable names are explicitly defined.

More information about data verbs and data grouping can be found in
[Chapter 5](https://r4ds.had.co.nz/transform.html) of the R for Data Science
textbook and in
[Notebook 4](https://statsmaths.github.io/stat289-f20/notebooks/notebook04.html),
[Notebook 5](https://statsmaths.github.io/stat289-f20/notebooks/notebook05.html),
and
[Notebook 6](https://statsmaths.github.io/stat289-f20/notebooks/notebook06.html)
of my Introduction to Data Science course.


### Combining Data: Two-table Verbs

All of the data verbs above work by taking a single data table as and input
and returning a modified copy of the data as an output. The other class of
data verbs that we will use allow us to combine information from two different
data tables. These are called *two-table verbs*. To show an example of these,
let's load another dataset of recipes showing the ingredients for two dishes.

```{r, message=FALSE}
recipes <- read_csv(file.path("data", "food_recipes.csv"))
recipes
```

We might want to combine this data with the food data to see, for example, how
many calories are in each dish. We do this with the `inner_join` function, which
allows us to combine two data sets by joining along a common key variable (here,
the food name).

```{r}
recipes %>%
  inner_join(food, by = c("ingredient" = "item"))
```

All of the food nutritional facts are given for a 100g serving; the recipes
give amounts in grams. With this knowledge, we can put together the verbs from
the previous sections to compute the amount of calories in each dish:

```{r}
recipes %>%
  inner_join(food, by = c("ingredient" = "item")) %>%
  mutate(calories_total = (calories / 100) * amount) %>%
  group_by(recipe) %>%
  summarize(sm_sum(calories_total))
```

Another two-table verb is `left_join`, which works exactly the same but includes
rows that only exist in the first table. Notice the difference here, with the
row containing the bay leaf (which is not present in the food dataset) included
in the output:

```{r}
recipes %>%
  left_join(food, by = c("ingredient" = "item"))
```

There are also the variations `right_join` and `full_join` which include
missing keys in the second data set and missing keys in both data sets,
respectively. Finally, the functions `semi_join` and `anti_join` identify rows
that have matching values in the two data sets, but do not actually perform
any joining together of the data columns. These will be useful in some text
analysis tasks.

More information about combining data sets based on common key variables can
be found in
[Chapter 12](https://r4ds.had.co.nz/tidy-data.html) and
[Chapter 13](https://r4ds.had.co.nz/relational-data.html)
of the R for Data Science
textbook and in
[Notebook 9](https://statsmaths.github.io/stat289-f20/notebooks/notebook09.html)
of my Introduction to Data Science course.

## Visualization

### Scatterplots

Another major task in data analysis is producing visualizations of data. For
this, we will use a system called the Grammar of Graphics. It requires a bit
of work to create simple plots, but can be extending in a logical way to capture
almost any kind of plot you would want to make with your data.

To start, let's see how to draw a scatter plot of our food data. Each row of
the data will be draw as a dot, with the x-coordinate given by the sugar content
of the food and the y-axis given by the number of calories in the food item.
This requires specifying three elements in the grammar of graphics:

- the data set (`food`)
- the type of *geometry* (points: `geom_point`)
- the x and y *aesthetics* (x: `sugar`, y: `calories`)

The syntax for doing this in R is:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = sugar, y = calories))
```

We can specify additional aesthetics that describe the way the points are
plotted by mapping these to other variables in the data. R will take care of
the details for us. For example, we can specify that the color of the points
should change based on the item's food group:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = sugar, y = calories, color = food_group))
```

Notice that R has figured out what colors to use and how to map them to each
unique value of the food group variable. Aesthetics can also be assigned to
different fixed values as follows (note that these arguments go outside of the
`aes()` function):

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = sugar, y = calories), color = "salmon", size = 4)
```

Resources for additional geometry types, aesthetics, and ways of further
customizing graphics are given at the end of the following section.

### Layering Graphics

In order to make more complex plots, we can layer multiple geometries together
by literally adding them together with the plus sign. For example, we can add
a text-repel layer to the plot that labels some of the items with labels (the
term repel indicates that the label will be made to avoid intersecting the point
and other labels). This geometry requires specifying the *label* aesthetic to
indicate which variable is used to provide the label.

```{r, warning = FALSE}
food %>%
  ggplot() +
    geom_point(aes(x = sugar, y = calories)) +
    geom_text_repel(aes(x = sugar, y = calories, label = item))
```

We can further modify the plot by creating a dataset of just the fruits and
labeling only the fruits (but including points for all of the layers). We do
this by manually specifying the data argument to the geometry layer. We will
also simplify the code by specifying the x- and y-aesthetics once in the
`ggplot()` function (it becomes the default for subsequent layers).

```{r, warning = FALSE}
fruit <- food %>%
  filter(food_group == "fruit")

food %>%
  ggplot(aes(x = sugar, y = calories)) +
    geom_point() +
    geom_text_repel(aes(label = item), data = fruit)
```

A more complete plot can be created by layering two sets of points and a
text-repel layer along with custom color names.

```{r, warning = FALSE}
food %>%
  ggplot(aes(x = sugar, y = calories)) +
    geom_point(color = "grey85") +
    geom_point(color = "navy", data = fruit) +
    geom_text_repel(aes(label = item), color = "navy", data = fruit)
```

This begins to show how relatively simple elements can be combined to produce
a nearly limitless set of increasingly complex data visualiations that help
explore the structure of our data.

More information about data visualization in R can be found in
[Chapter 3](https://r4ds.had.co.nz/data-visualisation.html)
of the R for Data Science textbook and in
[Notebook 2](https://statsmaths.github.io/stat289-f20/notebooks/notebook02.html),
[Notebook 3](https://statsmaths.github.io/stat289-f20/notebooks/notebook03.html),
and
[Notebook 8](https://statsmaths.github.io/stat289-f20/notebooks/notebook08.html)
of my Introduction to Data Science course.

## Additional Notes

### Coding Style Guidelines

It is very important to properly format your code in a consistent way.
Even though the code may run without errors and produce the desired results,
you will make your life easier by writing well-formatted code from the start.
This makes it easier to read and debug in the future. We will follow the
following guidelines:

- always put one space before and after an equals sign or assignment arrow
- always put one space around mathematical operations (such as `+` and `*`)
- always put one space after a comma, but no space before a comma
- variable names should only use lower-case letters, numbers, and underscores
- indent every line that follows a pipe with two extra spaces
- indent every line below `ggplot` by an additional two spaces (four total)

As with our original set of style guidelines, you will make your life a lot
easier if you get used to these rules right from the start.

### Continued Study

The material in these notes gives an overview of topics that are usually covered
over 4-5 weeks in MATH209 and/or MATH289. I do not expect anyone to completely
understand the implications and mechanics of these techniques on a first pass.
If the material is new to you, I suggest returning to these notes frequently
through the first few weeks of the semester as we encounter these methods in
our analysis of large textual corpora. As time allows, I suggest also following
the links to the course materials and the R for Data Science textbook.
