---
title: "Notebook 09: Documents as Vectors"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r message=FALSE}
library(tidyverse)
library(forcats)
library(ggrepel)
library(smodels)
library(cleanNLP)
library(Matrix)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
```

## Food Data

Today we make a shift towards a different type of textual analysis. So far we
have been mostly interested in a large collection of texts with the goal of
associating features in predicting some response variable. Now, we start to
consider the case where we are interested in understanding specific documents
in a corpus. Typically this will involve a smaller collection of longer
documents. To illustrate this, we will look at a data set of Wikipedia pages
corresponding to 61 food items (the same as in the example data from 289, if you
were in that class). We will load the corpus and the associated tokens. This
time, we will not worry about spliting the data into training and validation
sets.

```{r, message=FALSE}
food <- read_csv("data/food_page.csv")
token <- read_csv("data/food_page_token.csv.gz")

food
```

Notice that the document ids are a bit more informative that usual, and indicate
the name of the Wikipedia page in question.

## TF-IDF

We have already made extensive use of the term frequency matrix in our
predictive models. Here, we are going to consider a variation of this data
structure that includes some scaling of the values that can be useful for
exploring a corpus in a different way. It will be useful to have a way to
work with the term frequencies in a data frame, with one value in each
row, as follows:

```{r}
token %>%
  sm_text_tfidf(token_var = "lemma", min_df = 0, max_df = 1)
```

Note that regardless of what variable is used as `token_var`, the column
containing the thing being counted will be called `token`. The count variable is
stored in a column called `tf`, which stands for "term frequency". There is
another term called `tfidf` (TF-IDF), the term frequency-inverse document
frequency score. It takes a (scaled version) of the term frequency and divides
by (a scaled) proportion of documents that use the term. Mathematically, if
`tf` are the number of times a term is used in a document and `df` are the
proportion of documents that use the term at least once, the TF-IDF score can
be computed as:

$$ \text{tfidf} = \frac{(1 + log_2(\text{tf}))}{log_2(\text{df})} $$

The score gives a measurement of how important a term is in describing a
document in the context of the other documents. Note that this is a popular
choice for the scaling functions, but they are not universal and other software
way use different choices.

We can use TF-IDF to try to measure the most important words in each document.
Here, we filter out particularly rare terms that occur in less than 10% of the
documents, and then find the 8 nouns that have the highest value of `tfidf` for
each document:

```{r}
token %>%
  filter(upos == "NOUN") %>%
  sm_text_tfidf(min_df = 0.1) %>%
  arrange(desc(tfidf)) %>%
  group_by(doc_id) %>%
  slice(1:8) %>%
  summarize(sm_paste(token))
```

Do these capture words that best describe each page? How would you expect the
top terms to change if we applied it to a larger collection of Wikipedia pages
that included many non-food related articles?

## Documents as Vectors: Illustration

This section illustrates a concept that will be very useful in the final two
sections of this chapter. Note, however, that there is generally no need to
include this in your own analysis of textual data.

The TF-IDF data set is an example of a long-format data set. Conceptually, we
can think about the idea of converting this into a wide-format. Here, each row
would correspond to a document; variables would exist for each unique token,
giving counts corresponding to each document. This object can get quite large,
but writing the code is relatively straightforward using the techniques. Here,
we will filter to include only two lemmas, "animal" and "food", and pivot the
TF-IDF data set into a wide format.

```{r}
token %>%
  sm_text_tfidf() %>%
  filter(token %in% c("animal", "food")) %>%
  select(doc_id, token, tf) %>%
  pivot_wider(
    names_from = "token",
    values_from = "tf",
    names_prefix = "lemma_",
    values_fill = list("tf" = 0)
  )
```

Using just these two columns, we can plot a set of pages with `lemma_food` on
the x-axis and `lemma_animal`. It will be useful to think of these are vectors
starting at the origin, rather than points floating in space.

```{r}
token %>%
  sm_text_tfidf() %>%
  filter(token %in% c("animal", "food")) %>%
  select(doc_id, token, tf) %>%
  pivot_wider(
    names_from = "token",
    values_from = "tf",
    names_prefix = "lemma_",
    values_fill = list("tf" = 0)
  )  %>% filter(
    doc_id %in% c("Apple", "Beef", "Chicken", "Potato", "Milk", "Lamb", "Cheese")
  ) %>%
  ggplot() +
    geom_text(
      aes(x = lemma_food, y = lemma_animal, label = doc_id),
      nudge_x = 0.8,
      nudge_y = 0.8
    ) +
    geom_segment(
      aes(x = 0, y = 0, xend = lemma_food, yend = lemma_animal),
      arrow = arrow(length = unit(0.3,"cm"))
    )
```

What you should notice from this diagram is that these two words do a good job
of distinguishing the various pages. Beef and Lamb refer to animal food
products, and therefore have the highest usage of the lemma "animal". Potato
and Apple are not related to animals are all, and only use the lemma "food".
Milk and Cheese are food derived from animal products and sit in the middle of
the plot. Chicken is an animal, but its page focuses heavily on its culinary
usage, and therefore it sits closer to the dairy products.

The take-away from this illustration is that a wider-format of the term
frequency values provides an interesting way of grouping and exploring the
relationships between documents. Generally, we do not want to actually use
the `pivot_wider` function because it is too slow and clunky to work with a
data set that may have thousands of columns. Instead, we will use a different
approach with allows us to think of documents as living in a high dimensional
space without having to work with these large dimensional spaces directly.

## Dimension Reduction

Consider extending the illustration in the previous section to include a larger
set of lemmas. While we do not have an easy way of plotting the concept, we can
try to imagine each document as a an arrow from the origin to a point in a very
high dimensional space (one dimension for each unique token in the term
frequency dataset). In this section we will see a way of trying to work with
this high-dimensional space.

It was mentioned above that the `pivot_wider` function is not a good choice for
making a wider version of a term frequency data set. A better choice is the
function `cnlp_utils_tf`, provided by **cleanNLP**, which we have used
extensively so far in this course.

Principal component analysis is a common method for taking a high-dimensional
data set and converting it into a smaller set of dimensions that capture many
of the most interesting aspects of the higher dimensional space. The first
principal components is defined as a direction in the high-dimensional space
that captures the most variation in the inputs. The second component is a
dimension perpendicular to the first that captures the highest amount of
residual variance. Additional components are defined similarly. We can
compute principal components from the "dgCMatrix" using the helper function
`sm_tidy_pca`:

```{r}
token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_pca(n = 2)
```

The specific mathematics behind the principal components is less important than
how we interpret the output. Generally, went we plot the first 2 or 3 components
together, we do not worry about the specific dimensions. Rather, we want to use
the principal components to show relationships between documents based on
clusters and other information. Here is a plot of our data set on the
first two principal components:

```{r, warning = FALSE}
token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_pca() %>%
  ggplot(aes(x = v1, y = v2)) +
    geom_point(color = "grey90") +
    geom_text_repel(
      aes(label = document),
      show.legend = FALSE
    ) +
    theme_void()
```

Notice that the seafood and fruits cluster along different parts of the plot.
The other food groups roughly cluster in the upper-left hand side of the plot,
with vegetables closer to the fruits. The Duck page, an aquatic bird, is sit
in-between the seafood and other meats. All of these capture general
relationships we might expect given our knowledge of the pages and what they
represent.

Another method for reducing the dimension of our dataset is called UMAP
(Uniform Manifold Approximation and Projection). It has a much more complex
algorithm that is able to better spread the dataset uniformly over the plot
region. We can run this algorithm using the function `sm_tidy_umap`:

```{r, warning=FALSE}
token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_umap() %>%
  ggplot(aes(x = v1, y = v2)) +
    geom_point(color = "grey90") +
    geom_text_repel(
      aes(label = document),
      show.legend = FALSE
    ) +
    theme_void()
```

As with the principal components, the exact dimensions are unimportant here,
its the relationship between the documents that counts. Notice that the pages
are less clumped together here, but also that the structures from the principal
component analysis are not as clearly defined. The benefits of UMAP become more
apparent with larger datasets.

## Document Distance

In the previous two sections we treated documents as being points in a
high-dimensional space, and commented that our primary objective is
understanding the relationships between the documents in this space. We can
approach this question more directly by computing the distances between
documents in the high-dimensional space. This can be done by using the
`sm_tidy_distance` function applied to the TF-IDF matrix:

```{r}
token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_distance()
```

The output gives the distance between every pair of documents. Self-pairs (the
distance between Apple and Apple) are included to assist with other kinds of
analysis. Using this output, we can join each page to its closest neighbor:

```{r}
token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_distance() %>%
  filter(document1 != document2) %>%
  group_by(document1) %>%
  arrange(distance) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  arrange(distance)
```

Unfortunately, these links are largely not very intuitive Penne and
Flounder, for example, do not have very much in common. The issue here is that
the length of each document has a strong influence on the distance between
points. Returning to our illustrative example with just two lemmas, notice
that Potato is actually closer to Milk than it is to Apple. As an alternative,
we can compute the *angle* between two vectors using the function
`sm_tidy_angle_distance`.

```{r}
token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_angle_distance() %>%
  filter(document1 < document2) %>%
  group_by(document1) %>%
  arrange(distance) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  arrange(distance)
```

These relationships seem much more as expected, linking for example Milk and
Yogurt, Potato and Sweet Potato, Broccoli and Cauliflower. We will do more with
these distances as a form of network analysis in a subsequent notebook.

## Word Relationships

In all of the preceding analyses, we have focused on the analysis of the
document their usage of words. There are often multiple ways of widening a
data set, each leading to different kinds of analysis. The term frequency
data set is no different. We could widen the data set by treating each row as a
term and each column as a document. It is possible to apply dimensionality
reduction and distance metrics on this format as well in order to understand
the relationships between words.

The easiest way to produce a  matrix of the word relationships is by first
using `cnlp_utils_tfidf` as before and then calling the function `t()`
(transpose) to exchange the rows and columns. We will control the maximum
number of features by setting `max_features` to 100 and only considering nouns.
Here is the principal component analysis plot:

```{r, warning=FALSE}
token %>%
  filter(upos == "NOUN") %>%
  cnlp_utils_tfidf(
    min_df = 0, max_df = 1, max_features = 100
  ) %>%
  t() %>%
  sm_tidy_pca(item_name = "word") %>%
  ggplot(aes(x = v1, y = v2)) +
    geom_point(color = "grey90") +
    geom_text_repel(
      aes(label = word),
      show.legend = FALSE
    ) +
    theme_void()
```

As well as the closest pairs of words (here we increase the number of words
to 400):

```{r, warning=FALSE}
token %>%
  filter(upos == "NOUN") %>%
  cnlp_utils_tfidf(
    min_df = 0, max_df = 1, max_features = 400
  ) %>%
  t() %>%
  sm_tidy_angle_distance(item_name = "word") %>%
  filter(word1 < word2) %>%
  group_by(word1) %>%
  arrange(distance) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  arrange(distance)
```

Do these relationships seem reasonable to you? Do they tell you anything about
the data or the usage of language within the data that you find surprising?
