\documentclass[12pt, a4paper]{article}

\usepackage{fontspec}
\usepackage{geometry}
\usepackage{lastpage}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage[normalem]{ulem}
\usepackage{soul}
\usepackage{multicol}
\usepackage[dvipsnames]{xcolor}

\geometry{
  top=1.5cm,
  bottom=1.5cm,
  left=2cm,
  right=2cm,
  marginparsep=4pt,
  marginparwidth=1cm
}

\definecolor{lgrey}{rgb}{0.8, 0.8, 0.8}

\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancyplain}
\fancyhf{}
% \lfoot{\color{lgrey}2020-12-01}
% \rfoot{\color{lgrey}page \thepage\ of \pageref{LastPage}}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}

\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}

\setromanfont{YaleNew}

\begin{document}

\begin{center}
\textbf{MATH/STAT 389: Statistical Learning, Spring 2021, ONLINE}
\end{center}

\noindent
\begin{tabular}{ l l }
\textbf{Instructor:} &  \textbf{Taylor Arnold} \\
E-mail: & \texttt{tarnold2@richmond.edu} \\
Website: & \texttt{https://statsmaths.github.io/stat389-f21} \\
Office Hours: & Following Class; By Appointment; 19h evenings before projects due
\end{tabular}

\vspace{0.5cm}

\textbf{Format:} \vspace{6pt}

The course will be taught remotely over Zoom. During the first week of the
semester students will be organize into groups of around 3 people each.
Students may self-select their own groups. A large portion of the course work
will be done using break-out groups within these teams. Materials and
assignments for the class will be posted on the course website. All assignments
will be submitted through shared Google Drive folders.

\vspace{12pt}

\textbf{Projects:} \vspace{6pt}

The core output of the course will be the completion of five projects. These
will be assigned throughout the course of the semester and focus on the
application of the course material to various textual corpora. Projects will
be completed together in your class group. There will usually be plenty of
class time to work on each project. We will also have time to present a subset
of the projects to the rest of the class. Tentative due-dates and topics are
posted on the course website and will be updated as needed.

\vspace{12pt}

\textbf{Class Labs:} \vspace{6pt}

Other than when working or presenting the class projects, each class will have
a `lab' assigned to it. This is a set of programming and analysis questions
posed as R markdown files. Teams will work together in class by picking a
`driver' (which should alternate between classes) to share their screen and
take notes or write/run code. Groups should upload the version of the lab up to
wherever they have stopped in class showing their progress, but no formal grade
will be given.

\vspace{12pt}

\textbf{Attendence:} \vspace{6pt}

Due to the large amount of group-work required during the course meetings, it is
important for everyone to make an attempt to attend every class. With the
exception of extended absenses due to serious illness or family emergencies,
students will recieve a reduction in their final grade (including the
possibility of a failing grade) after missing more than two class meetings.

\vspace{12pt}

\textbf{Grades:} \vspace{6pt}

Projects will be graded as either Satisfactory/Unsatisfactory. At the end of
the semester, each student will individually submit a self-assessment of their
work in the course, how they contributed to each project, and what they have
learned during the semester.\\

All students who have earned satisfactory grades on the five projects, missed
no more than two classes, and have completed the course labs will earn a grade
of at least a B+. Grades of B+, A-, A, and A+ will be given according to each
student's self-assessment in light of their projects and course participation.

\vspace{12pt}

\textbf{Getting Help:} \vspace{6pt}

Given social distancing rules, I find the easiest way to answer questions
is by finding time after our scheduled class meeting times. I will always make
a habit of staying after class as long as students have questions. There will
also be office hours the evening before projects are due in case of last minute
questions. Finally, if you have extended questions, personal concerns, or are
unable to meet on either side of the class meeeting, please feel free to email
me or ask in class for a dedicated appointment.

\end{document}
