---
title: "Lab 05"
output: html_document
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(Matrix)
library(glmnet)
library(cleanNLP)
library(magrittr)

theme_set(theme_minimal())
options(sparse.colnames = TRUE)
options(dplyr.summarise.inform = FALSE)
```

# Chicago Crimes

## Three Types

Today's lab looks at a data set of reported crimes from the city of Chicago
(if you were in my 289 class last semester, this should look familiar). Our
prediction task is to predict the type of crime based on the features of the
reported incident. To start, we'll look at a data set that has three crime
types: "battery", "criminal damage", and "theft".

```{r, message=FALSE}
set.seed(1)

chicago <- read_csv(file.path("data", "chi_crimes_3.csv")) %>%
  mutate(train_id = if_else(runif(n()) < 0.6, "train", "valid"))
chicago
```

Start by creating a model matrix, response vector, and training model matrix
and response vector using all of the variables in the data set other than the
`train_id`. This will be fairly large due to the large number of options for
the variable `location`.

```{r, question-01}

```

Now, fit a cross-validated elastic net model using 3 folds and alpha equal to
0.9.

```{r, question-02}

```

Next, plot the model using the `plot` function. Take note of the shape of the
curve and the number of included variables (the numbers of the top of the plot)
for each value of lambda.

```{r, question-03}

```

And compute the classification rate for the data set.

```{r, question-04}

```

How does this compare to random guessing (the data are balanced, with an equal
number of crimes of each type)? **Answer**:

Now, build a confusion matrix of the data.

```{r, question-05}

```

Focusing on the validation set, are some crimes harder to distinguish than
others? **Answer**:

Now, print out the coefficients for a value of lambda that selects about a
dozen variables (perhaps try the 21st value of lambda).

```{r, question-06}

```

Looking at the selected variables, what locations seem to be most associated
with each type of crime? **Answer**:

Now, refit the model using alpha equal to 0.1.

```{r, question-07}

```

Now, plot the model.

```{r, question-08}

```

Verify that more non-zero components are present at the minimul value of the
CV-curve.

## Twelve Types

Let's now load another version of the crimes data, this time with 12 different
categories. The data set is large, so in the interest of time we will only use
5% of it for the training set to speed things along.

```{r, message=FALSE}
set.seed(1)

chicago12 <- read_csv(file.path("data", "chi_crimes_12.csv")) %>%
  mutate(train_id = if_else(runif(n()) < 0.05, "train", "valid"))
chicago12
```

Again, build a model matrix and response vector with all of the variables other
than `train_id`.

```{r, question-09}

```

We are going to again fit a cross-validated elastic net model with alpha 0.9,
and three folds. Add the option `trace.it = TRUE` to have a verbose print out
of the model progress (it may take a minute or two depending on your machine).

```{r, question-10}

```

Plot the model CV curve. Verify that the curve has a steeper bend for small
values of lambda compared to the previous model.

```{r, question-11}

```

Now, determine the classification rate.

```{r, question-12}

```

How does this compare to random guessing (the data are balanced, with an equal
number of crimes of each type)? **Answer**:

As before, produce a confusion matrix of the model:

```{r, question-13}

```

What are some crimes that seem difficult to tell apart on the validation set?
**Answer**:

For each (actual) crime type, compute the error rate for this category on the
validation set and arrange the data from the highest classification rate to
the lowest.

```{r, question-14}

```

What types are the hardest to get correct? **Answer**:

Finally, print out the coefficients for the model by picking a lambda that
gives about 20 non-zero terms. Take note of any patterns you see in the detected
locations.

```{r, question-15}

```

## Space and Time

Using the smaller Chicago data set with just three crimes, create a model matrix
with the following formula:

`crime_type ~ cut(longitude, breaks = 10) * cut(latitude, breaks = 10) * cut(hour, breaks = 3)`

This breaks space and time into individual buckets.

```{r, question-16}

```

Now, fit a cross-validate elastic net on the data using alpha = 0.9 and three
folds. This may take a minute or two to run, so consider turning on the
`traceit` flag.

```{r, question-17}

```

Finally, predict the class impied by the model for the data set and produce a
scatter plot with longitude on the x-axis, latitude on the y-axis, and the
points colored by the prediction. Also, add this facet wrap layer to produce
a different plot for each third of the day:
`facet_wrap(~cut(hour, breaks = 3), ncol = 4)`

```{r, question-18}

```

Take note of the patterns of Crime in different parts of the city for different
times of day. Notice how well our predictive model is able to describe patterns
that would be hard to find just exploring the source data itself.
