---
title: "Lab 01"
output: html_document
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(Matrix)
library(glmnet)
library(cleanNLP)
library(magrittr)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
```

This is the first "lab" for the semester. I have included several empty code
chunks and text prompts that start with "**Answer:**". You should fill in 
solutions to these (code in the first case and a short answer in the second).
These are not handed in or graded and solutions will be posted for next class.
I suggest generally working as a group with a single shared screen, but you
might want to also duplicate the code in your own machine.

## California House Prices

In this lab you will apply the methods shown in the notes to a data set of 
housing prices from California. Our variable of interest is the median
house sale price in each census tract.

```{r, message=FALSE}
set.seed(1)

ca <- read_csv(file.path("data", "ca_house_price.csv")) %>%
  mutate(train_id = if_else(runif(n()) < 0.6, "train", "valid"))
ca
```
There are several numeric variables that you can use to predict the output,
all of which should be reasonably self-explanatory.

### Exploring the Data

To start, draw a histogram with 15 bins showing the distribution of the
median house value variable:

```{r, question-01}
ca %>%
  ggplot(aes(median_house_value)) +
    geom_histogram(bins = 15, color = "black", fill = "white")
```

What are the most typical values of the variable (they are given in
thousands of dollars)? **Answer**: There are many values around \$300k dollars,
with a long tail ranging from near \$0 to just under \$1.

Now, draw a scatter plot showing the relationship between the median household
income and the median house value.

```{r, question-02}
ca %>%
  ggplot(aes(median_household_income, median_house_value)) +
    geom_point()
```

How would you describe this relationship? Is it surprising or as you would
expect? **Answer**: In general, areas with a higher median income have more
expensive houses, but the relationship has a lot of noise.

Now show the relationship between the mean household income and the median
house value.

```{r, question-03}
ca %>%
  ggplot(aes(mean_household_income, median_house_value)) +
    geom_point()
```

How does this compare to the previous plot? **Answer**: The relationship is
very similiar, and hard to differentiate just by a quick glance.

And now, show the relationship between the mean household income and
the median household income. Add a line with a slope of 1 with the following
layer: `geom_abline(slope = 1, color = "orange")`.

```{r, question-04}
ca %>%
  ggplot(aes(median_household_income, mean_household_income)) +
    geom_point() +
    geom_abline(slope = 1, color = "orange")
```

You should see that the mean and median income are highly correlated.
Which one tends to be slightly higher? Can you think of why the data are
distributed like this? **Answer**: The mean income ranges from right around
the median income (these are areas that likely have decent income equality)
to many times larger than this. These larger regions have a heavy tail with
a small number of higher earners pushing the mean income higher.

Finally, draw a scatter plot with longitude and latitude on the x- and y-axes.
Color the points according to the median house value and include the scale
`scale_color_viridis_c()`. You might also decrease the size to make the plot
easier to read.

```{r, question-05}
ca %>%
  ggplot(aes(longitude, latitude)) +
    geom_point(aes(color = median_house_value), size = 0.5) +
    scale_color_viridis_c()
```

See if you can identify the San Francisco, Los Angeles, and San Diego
metro regions. Where do you see the highest house values? How about the
lowest house values? **Answer**: The highest house values are in the three
large metro areas, particularly San Francisco. Lowest prices are inland.

### Simple Linear Models

Now, produce a linear model that predicts the median house value as a function
of the median household income. Make sure to use only the training data. Print
out a summary of the model.

```{r, question-06}
model <- ca %>%
  filter(train_id == "train") %>%
  lm(median_house_value ~ median_household_income, data = .)

summary(model)
```

Describe in words what the model thinks is the relationship between the
median household income and the median house value. **Answer**: The model
thinks that each extra dollar of median household income increases the typically
value of the house value by \$4.75.

Using the model you just created, compute the RMSE for the training and
validation sets.

```{r, question-07}
ca %>%
  mutate(value_pred_lm = predict(model, newdata = .)) %>%
  group_by(train_id) %>%
  summarize(rmse = sqrt(mean((median_house_value - value_pred_lm)^2)))
```

It's hard to see how good this prediction is without a reference point.
Compute the RMSE that comes from using the mean of the median house value
as a predictor.

```{r, question-08}
ca %>%
  mutate(value_pred_const = mean(median_house_value)) %>%
  group_by(train_id) %>%
  summarize(rmse = sqrt(mean((median_house_value - value_pred_const)^2)))
```

How does the RMSE of the linear model compare to the constant model? Does the
linear model seem to perform significantly better? **Answer**: Yes! The simple
model has an RMSE just under 200 whereas the linear regression is around 154.

To illustrate the linear model, draw a scatter plot with median household
income on the x-axis, median house value on the y-axis, and a line showing
the predicted linear model.

```{r, question-09}
ca %>%
  mutate(value_pred_lm = predict(model, newdata = .)) %>%
  ggplot(aes(median_household_income, median_house_value)) +
    geom_point() +
    geom_line(aes(y = value_pred_lm), color = "orange", size = 1.5)
```

Confirm that the linear seems to pass through the cloud of data in a reasonable
way.

Now, produce a linear regression model that uses the mean household income to
predict the median house value. Print out a summary of the model.

```{r, question-10}
model <- ca %>%
  filter(train_id == "train") %>%
  lm(median_house_value ~ mean_household_income, data = .)

summary(model)
```

How do the slope and intercept compare to the previous model? **Answer**: The
slope is actually quite close (4.5 instead of 4.75). The intercept is lower 
though, likely to account for the fact that the mean household income often
being larger than the median household income.

Now, compute the RMSE of this model.

```{r, question-11}
ca %>%
  mutate(value_pred_lm = predict(model, newdata = .)) %>%
  group_by(train_id) %>%
  summarize(rmse = sqrt(mean((median_house_value - value_pred_lm)^2)))
```

Which seems to be a more powerful estimator, the mean income or the median
income of each region? **Answer**: The mean income has an RMSE around 138, 
less than the previous model. It seems that mean income is a better predictor
than median income.

### Multivariate Linear Models

We will finish by producing multivariate linear regression models using more
than one predictor variable. Fit a regression model that uses both the mean
household income and the median household income to predict the house value.
Print out a summary of the model.

```{r, question-12}
model <- ca %>%
  filter(train_id == "train") %>%
  lm(median_house_value ~ mean_household_income + median_household_income, data = .)

summary(model)
```

What are the signs of the slope parameters? Do these surprise you?
**Answer**: The mean household income has a positive sign, but the median
household income has a negative sign. The phenomenon of a parameter switching
sign in a multivariate model is common when we have two highly-correlated
variables.

Compute the RMSE of this new model on the training and validation sets.

```{r, question-13}
ca %>%
  mutate(value_pred_lm = predict(model, newdata = .)) %>%
  group_by(train_id) %>%
  summarize(rmse = sqrt(mean((median_house_value - value_pred_lm)^2)))
```

How does this compare to the RMSE of the previous models? **Answer**: The 
RMSE decreases very slightly compared to the previous model, from 138.5 down
to 136.5.

Produce a scatter plot with median household income on the x-axis, mean
household income on the y-axis, and the points colored according the predicted
value from the linear model. Use the viridis color scale mentioned in the
previous question.

```{r, question-14}
ca %>%
  mutate(value_pred_lm = predict(model, newdata = .)) %>%
  ggplot(aes(median_household_income, mean_household_income)) +
    geom_point(aes(color = value_pred_lm)) +
    scale_color_viridis_c()
```

Can you see the effect of the two different coefficients that were found in
the linear regression model?

### Build Your Own

Before finishing, try to build a more predictive model using any other variables
that you think might be useful. Compute the RMSE of the training and validation
set and feel free to iterate a few times to get the best fit.

```{r, question-15}
model <- ca %>%
  filter(train_id == "train") %>%
  lm(median_house_value ~ mean_household_income +
                          median_household_income +
                          latitude +
                          longitude +
                          mean_household_size_renters +
                          built_1939_or_earlier +
                          bedrooms_1 +
                          bedrooms_2, data = .)

ca %>%
  mutate(value_pred_lm = predict(model, newdata = .)) %>%
  group_by(train_id) %>%
  summarize(rmse = sqrt(mean((median_house_value - value_pred_lm)^2)))
```

I was able to get a model with an RMSE just slightly above 100.
