---
title: "Lab 01"
output: html_document
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(Matrix)
library(glmnet)
library(cleanNLP)
library(magrittr)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
```

This is the first "lab" for the semester. I have included several empty code
chunks and text prompts that start with "**Answer:**". You should fill in
solutions to these (code in the first case and a short answer in the second).
These are not handed in or graded and solutions will be posted for next class.
I suggest generally working as a group with a single shared screen.

## California House Prices

In this lab you will apply the methods shown in the notes to a data set of
housing prices from California. Our variable of interest is the median
house sale price in each census tract.

```{r, message=FALSE}
set.seed(1)

ca <- read_csv(file.path("data", "ca_house_price.csv")) %>%
  mutate(train_id = if_else(runif(n()) < 0.6, "train", "valid"))
ca
```
There are several numeric variables that you can use to predict the output,
all of which should be reasonably self-explanatory.

### Exploring the Data

To start, draw a histogram with 15 bins showing the distribution of the
median house value variable:

```{r, question-01}

```

What are the most typical values of the variable (they are given in
thousands of dollars)? **Answer**:

Now, draw a scatter plot showing the relationship between the median household
income and the median house value.

```{r, question-02}

```

How would you describe this relationship? Is it surprising or as you would
expect? **Answer**:

Now show the relationship between the mean household income and the median
house value.

```{r, question-03}

```

How does this compare to the previous plot? **Answer**:

And now, show the relationship between the mean household income and
the median household income. Add a line with a slope of 1 with the following
layer: `geom_abline(slope = 1, color = "orange")`.

```{r, question-04}

```

You should see that the mean and median income are highly correlated.
Which one tends to be slightly higher? Can you think of why the data are
distributed like this? **Answer**:

Finally, draw a scatter plot with longitude and latitude on the x- and y-axes.
Color the points according to the median house value and include the scale
`scale_color_viridis_c()`. You might also decrease the size to make the plot
easier to read.

```{r, question-05}

```

See if you can identify the San Francisco, Los Angeles, and San Diego
metro regions. Where do you see the highest house values? How about the
lowest house values? **Answer**:

### Simple Linear Models

Now, produce a linear model that predicts the median house value as a function
of the median household income. Make sure to use only the training data. Print
out a summary of the model.

```{r, question-06}

```

Describe in words what the model thinks is the relationship between the
median household income and the median house value. **Answer**:

Using the model you just created, compute the RMSE for the training and
validation sets.

```{r, question-07}

```

It's hard to see how good this prediction is without a reference point.
Compute the RMSE that comes from using the mean of the median house value
as a predictor.

```{r, question-08}

```

How does the RMSE of the linear model compare to the constant model? Does the
linear model seem to perform significantly better? **Answer**:

To illustrate the linear model, draw a scatter plot with median household
income on the x-axis, median house value on the y-axis, and a line showing
the predicted linear model.

```{r, question-09}

```

Confirm that the linear seems to pass through the cloud of data in a reasonable
way.

Now, produce a linear regression model that uses the mean household income to
predict the median house value. Print out a summary of the model.

```{r, question-10}

```

How do the slope and intercept compare to the previous model? **Answer**:

Now, compute the RMSE of this model.

```{r, question-11}

```

Which seems to be a more powerful estimator, the mean income or the median
income of each region? **Answer**:

### Multivariate Linear Models

We will finish by producing multivariate linear regression models using more
than one predictor variable. Fit a regression model that uses both the mean
household income and the median household income to predict the house value.
Print out a summary of the model.

```{r, question-12}

```

What are the signs of the slope parameters? Do these surprise you?
**Answer**:

Compute the RMSE of this new model on the training and validation sets.

```{r, question-13}

```

How does this compare to the RMSE of the previous models? **Answer**:

Produce a scatter plot with median household income on the x-axis, mean
household income on the y-axis, and the points colored according the predicted
value from the linear model. Use the viridis color scale mentioned in the
previous question.

```{r, question-14}

```

Can you see the effect of the two different coefficients that were found in
the linear regression model?

### Build Your Own

Before finishing, try to build a more predictive model using any other variables
that you think might be useful. Compute the RMSE of the training and validation
set and feel free to iterate a few times to get the best fit.

```{r, question-15}

```

I was able to get a model with an RMSE just slightly above 100.
