---
title: "Project 04 --- Create Data"
output: html_document
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(Matrix)
library(glmnet)
library(cleanNLP)
library(magrittr)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)

source("wiki.R")
```

You should use this file to create your dataset, and then use the other file
to read in your data and run the analysis as with the other projects. This
will help avoid needing to create the data set each time you are working with
the project.

Here, grab all of the links from one (or more) starter pages and see how many
pages there are. Try to aim for somewhere between 100 and 1500 pages.

```{r, message=TRUE}
starter_page_name <- c("")

starter_page <- wiki_get_pages(starter_page_name, lang = "en")
starter_page$links <- stri_replace_all(starter_page$links, "%26", fixed = "&")
exp_pages <- wiki_expand_pages(starter_page)
wiki <- wiki_get_pages_text(exp_pages)

wiki <- tapply(wiki$text, wiki$page, paste, collapse = " ")
wiki <- tibble(doc_id = names(wiki), text = as.character(wiki))

print(nrow(wiki))
```

If you need to, after looking at the data, use the following code to remove
any pages that you do not want in your analysis.

```{r, message=FALSE}
remove_these <- c()
wiki <- wiki[!(wiki$doc_id %in% remove_these), ]
```

And then, run the udpipe annotator over your data. It will be a bit slower
and not quite as accurate, as the spaCy annotator, but does not require setting
up Python on your machine (a real pain for just this one assignment).

```{r}
cnlp_init_udpipe("english")
token <- cnlp_annotate(wiki, verbose = 1)$token
```

And finally, save the data set here:

```{r}
write_csv(wiki, file.path("data", "wiki_project04.csv"))
write_csv(token, file.path("data", "wiki_project04_token.csv.gz"))
```

You can then read back into R using the code in `project04.Rmd`. If you don't
find the results very interesting, or need to adjust anything, you can do it
here
